---
title: "Shot predictor"
author: "Augustin de Menibus"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Exploration:

```{r}
library(dplyr)
library(ggplot2)
library(rsample)
library(recipes)
library(caret)



NBAshots <- read.csv("training.csv.gz")

str(NBAshots)

summary(NBAshots)

## How many values are missing in each column:
sort(colSums(is.na(NBAshots)),decreasing = TRUE)

```

## Necessary Data Transformation: Part 1: Shot type and Game state

These transformations only look one row at a time, or one shot a time. They are not learning from other rows so there is no information leakage between folds and are safe to run before the split.


```{r}
## We will make a copy df so we have a baseline to compare our data transformation to:

NBAshots_train = NBAshots


## Convert strings into categories for our model to learn different relationships for different categories:
NBAshots_train$gamestate = factor(NBAshots_train$gamestate, ordered = FALSE )
## There are different game states, therefore game state has distinct categories with no rankings. That is why we turn it into factors so the model can develop relationship per category.

NBAshots_train$shottype = factor(NBAshots_train$shottype, ordered = FALSE)
## The logic is the same as with game state, we are making none ranked categories that will be fed to our model.

## We will now analyze the actual categories and the frequencies that we have in our data set from game state and shot type
table(NBAshots_train$gamestate, useNA = "ifany")
table(NBAshots_train$shottype, useNA = "ifany")

Shotmadepertype <- tapply(NBAshots_train$outcome == 1 , NBAshots_train$shottype, mean, na.rm =TRUE)
Shotmadepertype

Shotmadepergamestate <- tapply(NBAshots_train$outcome == 1 , NBAshots_train$gamestate, mean, 
                               na.rm = TRUE) 
Shotmadepergamestate

## We notice clear patterns, shots that are closer to the basket are more likely to go in, this makes sense as they are simpler shots that have a higher probability of going in.
```

## Necessary Data Transformation 2: Defender approach transformation:

```{r}

# Remove braces and split each shot's text at the commas
approach_text = gsub("[{}]", "", NBAshots_train$closestdefapproach)
approach_list = strsplit(approach_text, ",", fixed = TRUE)

# Convert each shot's four values into numbers
parse_distances = function(x) {
  if (length(x) == 1L && is.na(x)) return(rep(NA_real_, 4L))

  if (length(x) != 4L) {
    stop("A shot does not contain four approach values.")
  }

  x = trimws(x)
  missing_value = is.na(x) | x %in% c("", "NA", "NULL", "null")
  values = suppressWarnings(as.numeric(x))

  if (any(!missing_value & !is.finite(values))) {
    stop("An approach value is not a valid number.")
  }

  values[missing_value] = NA_real_
  values
}

# Process every shot and arrange the results into rows
approach_matrix = t(
  vapply(approach_list, parse_distances, numeric(4))
)

# Name the four columns by seconds before release
approach_names = c(
  "defdist_1s",
  "defdist_075s",
  "defdist_05s",
  "defdist_025s"
)

# Add the columns to working data
NBAshots_train[approach_names] = as.data.frame(approach_matrix)

NBAshots_train$closestdefapproach = NULL

```


## Distribution and Variance Exploration: 
### Making the plotting function
```{r}
## We will now examine interesting distributions:

shots <- NBAshots_train

makesummary = function(shots, ...) {
  shots %>%
    group_by(...) %>%
    summarise(
      attempts = n(),
      knownoutcomes = sum(!is.na(outcome)),
      makes = sum(outcome == 1, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      make_pct = if_else(
        knownoutcomes > 0,
        100 * makes / knownoutcomes,
        NA_real_
      )
    )
}

plot_rates = function(summarytable, variable, axis_label, plot_title) {

  ggplot(
    summarytable,
    aes(x = .data[[variable]], y = make_pct)
  ) +
    geom_col(fill = "firebrick", width = 0.7) +
    geom_text(
      aes(label = sprintf("%.1f%%", make_pct)),
      nudge_y = 2,
      vjust = 0,
      size = 3
    ) +
    scale_y_continuous(
      limits = c(0, 110),
      breaks = seq(0, 100, 25),
      labels = function(x) paste0(x, "%")
    ) +
    labs(
      title = plot_title,
      x = axis_label,
      y = "Make percentage",
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 35, hjust = 1)
    )
}
```

### Visualizations part 1: Shot type and Game context: 
```{r}
# 1. Make percentage by shot type
library(ggplot2)
type_summary = makesummary(shots, shottype)

print(type_summary)

print(
  plot_rates(
    type_summary,
    "shottype",
    "Shot type",
    "Make percentage by shot type"
  )
)


# 2. Make percentage by game state

state_summary = makesummary(shots, gamestate)

print(state_summary)

print(
  plot_rates(
    state_summary,
    "gamestate",
    "Game state",
    "Make percentage by game state"
  )
)


# 3. Make percentage for two versus three point attempts

# Add readable labels while preserving the original three column
shots$point_value = factor(
  shots$three,
  levels = c(0, 1),
  labels = c("2 point", "3 point")
)

three_summary = makesummary(shots, point_value)

print(three_summary)

print(
  plot_rates(
    three_summary,
    "point_value",
    "Point value",
    "Make percentage for two and three point attempts"
  )
)
```

### Interpretation: 
We can see clear trends for the probability of a shot going in and the type of shot taken. We previously touched on this trend before, the bar graph confirms our initial hypothesis. Furthermore, we see the difference in probability of success also depend on the state of the game and if the shot is 2 or 3 pointer. 

## Visualization part 2: Shot distance, location and time: 
```{r}
# Visual 4: Make percentage by distance from the basket

shots$distance_band = cut(
  shots$distance,
  breaks = c(0, 3, 8, 16, 24, 30, Inf),
  include.lowest = TRUE
)

distance_summary = makesummary(shots, distance_band)

print(distance_summary)

print(
  plot_rates(
    distance_summary,
    "distance_band",
    "Distance from basket in feet",
    "Make percentage by shot distance"
  )
)

# Visual 5: Make percentage by shot clock interval

shots$shotclock_band = cut(
  shots$shotclock,
  breaks = c(0, 4, 8, 12, 16, 20, 24),
  include.lowest = TRUE
)

clock_summary = makesummary(shots, shotclock_band)

print(clock_summary)

print(
  plot_rates(
    clock_summary,
    "shotclock_band",
    "Seconds remaining on the shot clock",
    "Make percentage by shot clock interval"
  )
)


```

### Interpretation: 
Firstly, the variable of distance has a clear negative relationship with the probability of a shot being scored. As we get further from the basket it becomes more difficult to make shots. Secondly, we see a positive relationship between the probability of a shot being scored and the time left on the shot clock, the more time is left the more shots have been scored. A possible explanation could be that players are less rushed and stressed when shooting early in the shot clock.

## Visualization Part 3: Defensive pressure and movement
```{r}
# Visual 6: Make percentage by number of contesters

shots$contester_group <- factor(
  shots$num_contesters,
  levels = 0:4
)

contester_summary = makesummary(shots, contester_group)

print(contester_summary)

visual6 = plot_rates(
  contester_summary,
  "contester_group",
  "Number of contesters",
  "Visual 6: Make percentage by number of contesters"
)

print(visual6)


# Visual 7: Make percentage by closest defender distance

shots$defender_band = cut(
  shots$closestdefdist,
  breaks = c(0, 2, 4, 6, 10, Inf),
  include.lowest = TRUE
)

defender_summary = makesummary(shots, defender_band)

print(defender_summary)

visual7 = plot_rates(
  defender_summary,
  "defender_band",
  "Closest defender distance in feet",
  "Visual 7: Make percentage by closest defender distance"
)

print(visual7)


```

### Interpretation
There is a clear negative relationship between make percentage and number of contesters. As the number of contesters increases, the make percentage decreases. This seems logical as more defensive players increase the opportunity of blocks or slight alterations to jump shot trajectory. The position of the closest defender doesn't have a clear impact on shooting percentages. Interestingly, when defenders are in between 2 and 4 feet from the offensive player, NBA shots have the highest percentage of going in.



## Model part 1: Data Split + Folds
```{r}
## Create five fold cross validation for our models:
predictors = c("num_contesters",
               "distcont1", "distcont2", "distcont3", "distcont4",
               "contested", "gamestate", "shottype",
               "distance","locationx", "locationy",
               "three", "dribblesbefore", "shotclock",
               "closestdefdist", "shooterspeed",
               "defdist_1s", "defdist_075s",
               "defdist_05s","defdist_025s",
               "month","season_id")
## We define the original predictors given to us

modeldata <- shots %>%
  select(outcome, all_of(predictors)) %>%
  filter(!is.na(outcome)) %>%
  mutate(
    across(where(is.factor), as.character),
    contested = as.integer(contested),
    three = as.integer(three),
    season_id = as.character(season_id),
    month = as.character(month),
    .stratum = factor(outcome)
  )

## We make our initial split reproducible
set.seed(12345)

outersplit <- initial_split(
  modeldata,
  prop = 0.8,
  strata = .stratum)

development <- training(outersplit)
holdoutdata <- testing(outersplit)

## Now we create validation folds within our development data
set.seed(678)
v=5 
folds = vfold_cv(
  development,
  v=v,
  strata = .stratum
)


```


## Model Part 2: Preprocessing 
```{r}

## We need to clean our contester columns data as it holds N/A values:
contestercolumns = c("distcont1", "distcont2", "distcont3", "distcont4")
defendercolumns = c("closestdefdist", approach_names, contestercolumns)
continuouscolumns = c("num_contesters", defendercolumns, "distance", "locationx",
                      "locationy", "dribblesbefore", "shotclock", "shooterspeed")
## We must ensure that outcome and identifiers are not used as predictors, if not our model won't be accurate:
stopifnot(!any(c("outcome",".stratum","shot_id")%in% predictors))

## We need to store our cleaned data for every fold: 

folddata = list()

for(i in seq_len(nrow(folds))) {
  trainingdata = analysis(folds$splits[[i]])
  validationdata = assessment(folds$splits[[i]])
  ## Select predictors:
  traininginputs = trainingdata[, predictors, drop = FALSE]
  validationinputs = validationdata[, predictors, drop = FALSE]
  ## We make sure the categories exist:
  stopifnot("tip" %in% traininginputs$shottype,
            "TRANSITION" %in% traininginputs$gamestate)
  ##Preprocessing
  nbarecipe = recipe(~ ., data = traininginputs) %>%
    ## Categories that were not observed
    step_novel(all_nominal_predictors()) %>%
    ## Handle missing categorical values
    step_unknown(all_nominal_predictors()) %>%
    ## Set our reference categories:
    step_relevel(shottype, ref_level = "dunk") %>%
    step_relevel(gamestate, ref_level = "HALFCOURT") %>%
    ## This releveling was a deliberate choice
      step_indicate_na(all_of(defendercolumns), prefix = "na_ind") %>%
    step_mutate(across(all_of(defendercolumns), ~ coalesce(.x, distance))) %>%
    ## We choose to substitute N/A values with the distance of the actual shot
    step_impute_median(all_numeric_predictors()) %>%
    step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
    ## Remove columns that don't have any variance as they don't help our predictions
    step_zv(all_predictors()) %>%
    # Temporarily represent the intercept using a column of ones
    step_mutate(glm_constant = 1) %>%

# Identify and remove exactly redundant columns
   step_lincomb(glm_constant, all_numeric_predictors()) %>%
  step_rm(glm_constant) %>%
  step_normalize(any_of(continuouscolumns))

  ## Learn from the preprocessing
  preprocessor = prep(nbarecipe, training = traininginputs, retain = FALSE)
  ## We need to apply rules:
  xtrain = bake(preprocessor, new_data = traininginputs, composition = "matrix")
  xvalid = bake(preprocessor, new_data = validationinputs, composition = "matrix")
 
   ## Make sure we don't have missing values or infinite values that would create an error:
  stopifnot( ncol(xtrain) > 0,
             all(is.finite(xtrain)),
             all(is.finite(xvalid)),
             identical(colnames(xtrain),colnames(xvalid)))
  
  folddata[[i]] = list(
    xtrain = xtrain,
    ytrain = trainingdata$outcome,
    xvalid = xvalid,
    yvalid = validationdata$outcome,
    preprocessor = preprocessor
  )
  
}


```

## Basic Logistic Regression + log loss:

```{r}
## We are making a function to give a log loss score to our models:
logloss <- function(y, p, eps = 1e-15)
{
  p = pmin(pmax(p,eps), 1-eps)
  -mean(y+log(p) + (1-y)*log(1-p))
}

## Now we pit our model against a baseline:
oof_glm = rep(NA_real_,
              nrow(development))
              fold_loss_glm =
                numeric(length(folddata))
              
for (i in seq_along(folddata)) {
 trainingglm = 
   as.data.frame(folddata[[i]]$xtrain)
 trainingglm$outcome =
   folddata[[i]]$ytrain
 validationglm = 
   as.data.frame(folddata[[i]]$xvalid)

 simplemodel = glm(outcome ~ ., data = trainingglm,
                   family = binomial(link = "logit"))
 shotprobability = predict(simplemodel, newdata = validationglm,
                           type = "response")
 validationrows = complement(folds$splits[[i]])
 oofglm[validationrows] = shotprobability
 fold_loss_glm[i] = logloss(folddata[[i]]$yvalid, shotprobability)
}

              
stopifnot(!anyNA(oof_glm))
## We map the log loss per fold
fold_loss_glm
## Overall log loss of our model
logloss(development$outcome,oof_glm)

## The base line, which predicts the mean for every shot: 
baselineprobability = rep(mean(development$outcome), nrow(development))
logloss(development$outcome, baselineprobability)

```

